---
title: Download source data
format: html
---

<span style='color:red'>delete before publishing</span>

In a reproducible repository, we want all raw data to be stored on an [OGC Compliant](https://www.ogc.org/standards/) and [FAIR (Findable, Accessible, Interoperable, Reusable) Compliant](https://www.go-fair.org/fair-principles/) online geodatabases. You will use this code to reference these stable data download url's. This does two things:

The Geoportti project has curated a list of available geospatial and earth systems observations and model outputs. Check here to learn about types of data available: [Geoportti Geodata and Metadata](https://www.geoportti.fi/services/data/).

In Digital Waters, we will utilize the AquaINFRA DDAS for federated searches across dozens of public online geodatabases.

Your personal research data, including private or sensitive data, can be stored on [Zenodo](https://zenodo.org/) with a stable digital object identifier (stable download url). Please discuss this with your PI.

Data can also be stored and accessed on institutional networks (e.g. ftp or https servers), cloud storage accounts (Google Cloud, Amazon AWS, Azure), or on the [CSC.fi](CSC.fi) infrastructure.

# SE1_data_access
<span style='color:goldenrod'>modify before publishing</span> 

All raw source data are **not committed to the repository** to:
- keep the repo lightweight.
- prevent large file versioning.
- maintain reproducibility.
- separate raw vs processed data.

The raw data is stored in the subfolders of `inputs`.

```{r}
#| label: "common"
#| include: FALSE
source("_common.R") # Knitr hook to enable global R-env when rendering many .qmd
```

```{r}
here::i_am("SE1_data_access.qmd")
root = here::here()

raw_data_dir = fs::path(root, "inputs") # Define raw data dir var

fs::dir_create(raw_data_dir, c("manual","auto")) # Ensure folder structure

if (length(fs::dir_ls(fs::path(raw_data_dir, "manual"))) < 4L) { # TODO: set precise number of required manual dls
    stop("Datasets listed in README to require manual inputs are not found")
}
```

<span style='color:goldenrod'>modify before publishing</span> 
## Data description
Describe your data sources here. Reference the table provided on your README.

## How to run this notebook
Provide a detailed description of inputs, input directories, outputs, and output directories.

## Storage requirements
Total required storage: what is the total volume of data that will be downloaded with this notebook?
Maximum file size: what is the largest individual file size that will be downloaded with this notebook?

### Make sure you have sufficient storage available to download all data required
```{r}
if ((ps::ps_disk_usage()$available / 2^30) < 100) { # Gb TODO: precise storage req of auto datasets
    stop("Insufficient disk space for the automatically downloaded datasets")
}
```

<span style='color:red'>delete before publishing</span>

# Methods to automate data access 
Accessing data automatically using URLs (uniform resource locators, also known as web addresses) is fundamental to FAIR research (Findable, Accessible, Interoperable, Reproducible), in that it allows for others to access your data without creating redundant copies of files. We are particularly interested in using persistent identfiers (PIDs), which are URLs that do not change over time. One type of PID that we are very familiar with in research is a DOI (Digital Object Identifier). These are assigned to our publications, for example. 

There are three options to specify download URLs in this code:
1) List stable URLs in a "datalinks.txt" file in your "inputs" folder, and then iterate over these to download them. This option is ideal for smaller files that do not require user authentication to access.
2) Generate list of stable URLs using database intuition: this is ideal for ftp and http servers, where databasing syntax is easily inferred from the file URLs, without need for an API (advanced programming interface) to modify the search. Here, we can use regular expressions to guess the URLs for data representing different locations, points in time, or variables, or customize URLs to include details like our own username and password.
3) Accessing data though an API (Advanced Programming Interface): APIs allow us to use code to interact with database search parameters, find relavant data, and access download links for that relevant data. APIs help with searching complex, heterogenous geodatabases. For very large datasets (reanalysis data, satellite remote sensing data, etc), certain APIs allow us to subset data (say for a specific location) prior to downloading, reducing storage requirements. APIs often require environmnetal configuration for access control. If so, **access configuration details must be specified in README**

## Handling generated or simulated data
For experiments relying on synthetic data, please include synthetic data generation script in this notebook. Be sure to use features like set seed to ensure consistency with random number generation. Save all outputs to rawdata folder.


<span style='color:red'>delete before publishing</span>

# Define download file helper functions
This function will download files from URLs to the raw data folder created in the parent folder of the current repository. 
```{r helpers-for-download}
get_filename_from_url <- function(url) {
    httr2::url_parse(url)$path |> basename()
}
download_file <- function(url, out_dir) {
    old_o <-  options(timeout = max(300, getOption("timeout")))
    filename <- get_filename_from_url(url)

    filepath <- fs::path(out_dir, filename)
    if (fs::file_exists(filepath)) {
        print("File already exists, skipping: ", filename)
        options(old_o); return(TRUE)
    } else {
        print("Downloading: ", filename)
        req <- httr2::request(url) |>
            httr2::req_retry(max_tries=3, max_seconds=TIMEOUT, retry_on_failure=T,backoff=\(x){2}) |>
            httr2::req_progress()
        httr2::req_perform(req, path = filepath)
        print("✅ Saved: ", filepath)
        options(old_o); return(TRUE)
    }
}
```

```{r load-manual-data}
G <- data.table::fread(fs::path(raw_data_dir,"manual/G_reg.csv"), keepLeadingZeros = T)
stations <- data.table::fread(fs::path(raw_data_dir,"manual/asemat.csv"), keepLeadingZeros = T)
gwas <- data.table::fread(fs::path(raw_data_dir,"manual/pv_alueet.csv"), keepLeadingZeros = T)
pipes <- data.table::fread(fs::path(raw_data_dir,"manual/paikat.csv"), keepLeadingZeros = T)
```

```{r}
# ===== USER SETTINGS FOR SPI & SPEI DATA =====
OUTPUT_DIR <- fs::dir_create(raw_data_dir,"auto","ecmwfr_spi_spei_monthly")
TIMEOUT <- 2000 # Seconds before req timeout
# =========================
```

```{r download-spi-spei}
tryCatch(invisible(ecmwfr::wf_get_key()), error=\(e){ # Check for ECMWFR API key
    warning("Add European Centre for Medium-Range Weather Forecasts API key (ecmwfr::wf_set_key())")
    ecmwfr::wf_set_key()
    })
# For more on ecmwfr datasets: ecmwfr::wf_dataset_info(), ecmwfr::wf_datasets()

ecmwf_spi_spei_req_year <- function(y) {
    list(dataset_short_name = "derived-drought-historical-monthly",
        variable = c("standardised_precipitation_index",
                    "standardised_precipitation_evapotranspiration_index"),
        accumulation_period = c("1","3","6","12","24","36","48"),
        version = "1_0", product_type = "reanalysis",
        dataset_type = "consolidated_dataset",
        year = as.character(y),
        month = c("01","02","03","04","05","06","07","08","09","10","11","12"),
        area = c(70, 18, 59, 32), # Nmax, Emin, Nmin, Emax, WGS84
        target = paste0("ecmwfr_spi_spei_",y,".nc")  # Output filename & format
        # ("target" potentially converts to zip if the req would return many files)
    )
}
# test_dl <- ecmwfr::wf_request(ecmwf_spi_spei_req_year(1966), path=fs::dir_create(raw_data_dir,"auto","test"), time_out = 1000)

out_dir_dls <- fs::dir_create(OUTPUT_DIR, "downloads") # Dir for dls from API
out_dir_final <- fs::dir_create(OUTPUT_DIR, "unzipped") # Dir for final usable files

all_ecmwf_reqs <- lapply(1965:2024, ecmwf_spi_spei_req_year)

# Drop the API calls which files already exists
wanted_req_files <- fs::path_ext_remove(sapply(all_ecmwf_reqs, \(req) req$target))
existing_req_files <- fs::path_ext_remove(fs::path_file(fs::dir_ls(out_dir_dls, type="file")))
needed_ecmwf_reqs <- all_ecmwf_reqs[which(!(wanted_req_files %in% existing_req_files))]

if (!rlang::is_empty(needed_ecmwf_reqs)) {
    message("Downloading monthly SPI and SPEI...")
    message("This can take around half an hour")
    sent_reqs <- ecmwfr::wf_request_batch(
        request_list = needed_ecmwf_reqs,
        workers = min(20, length(needed_ecmwf_reqs)), # Max 20?
        time_out = TIMEOUT, #s
        retry = 20, # sec to wait between api calls
        path = out_dir_dls # Into subfolder of downloads
    )
    # Unzip all downloads into subfolder
    all_dl_zips <- fs::dir_ls(out_dir_dls, type="file", glob="*.zip")
    stopifnot(all(fs::is_file(all_dl_zips)))
    for (f in all_dl_zips) { unzip(f, exdir = out_dir_final) }
    rm(all_dl_zips, sent_reqs)
    # Possible to download also quality criteria for SPI & SPEI
    # Similar req to above, but variables "probability_of_zero_precipitation_spi", "test_for_normality_spi", "test_for_normality_spei"
    # All three cover all years in the dataset (no year needed in req)
    message("SPI and SPEI datasets downloaded.")
} else { message("All requested SPI & SPEI original files found") }
rm(wanted_req_files, existing_req_files, needed_ecmwf_reqs, all_ecmwf_reqs, out_dir_dls)
```

Download FMI's daily gridded data from around all pipes where natural QC'd GW data exists.
Data DOI reference: https://doi.org/10.35614/ISSN-2341-6408-IK-2022-11-RL
AWS S3 link assumed to be stable enough
[DRAFT, likely will be deleted]
