---
title: Untitled
format: html
---

<span style='color:red'>delete before publishing</span>

In a reproducible repository, we want all raw data to be stored on an [OGC Compliant](https://www.ogc.org/standards/) and [FAIR (Findable, Accessible, Interoperable, Reusable) Compliant](https://www.go-fair.org/fair-principles/) online geodatabases. You will use this code to reference these stable data download url's. This does two things:

The Geoportti project has curated a list of available geospatial and earth systems observations and model outputs. Check here to learn about types of data available: [Geoportti Geodata and Metadata](https://www.geoportti.fi/services/data/).

In Digital Waters, we will utilize the AquaINFRA DDAS for federated searches across dozens of public online geodatabases. Stay tuned for more information on how to search for data within the DIWA DataLab.

Your personal research data, including private or sensitive data, can be stored on [Zenodo](https://zenodo.org/) with a stable digital object identifier (stable download url). Please discuss this with your PI.

Data can also be stored and accessed on institutional networks (e.g. ftp or https servers), cloud storage accounts (Google Cloud, Amazon AWS, Azure), or on the [CSC.fi](CSC.fi) infrastructure.

# SE1_data_access
<span style='color:goldenrod'>modify before publishing</span> 

All raw source data are **not committed to the repository** to:
- keep the repo lightweight.
- prevent large file versioning.
- maintain reproducibility.
- separate raw vs processed data.

The raw data is stored in the subfolders of `inputs`.

```{r}
here::i_am("SE1_data_access.qmd")
root = here::here()

fs::dir_create(c(root, "inputs"), c("manual","auto")) # Ensure folder structure

raw_data_dir = fs::path(root, "inputs") # Define raw data dir var

if (length(fs::dir_ls(fs::path(raw_data_dir, "manual"))) < 4L) { # TODO: set precise number
    stop("Datasets listed in README to require manual inputs are not found")
}
```

<span style='color:goldenrod'>modify before publishing</span> 
## Data description
Describe your data sources here. Reference the table provided on your README.

## How to run this notebook
Provide a detailed description of inputs, input directories, outputs, and output directories.

## Storage requirements
Total required storage: what is the total volume of data that will be downloaded with this notebook?
Maximum file size: what is the largest individual file size that will be downloaded with this notebook?

### Make sure you have sufficient storage available to download all data required

```{python}
!df -k ..
# the table below prints out how much storate is available in the folder containing your raw_data download directory
```
```{r}
if ((ps::ps_disk_usage()$available / 2^30) < 100) { # Gb TODO: precise storage req of auto datasets
    stop("Insufficient disk space for the automatically downloaded datasets")
}
```

<span style='color:red'>delete before publishing</span>

# Methods to automate data access 
Accessing data automatically using URLs (uniform resource locators, also known as web addresses) is fundamental to FAIR research (Findable, Accessible, Interoperable, Reproducible), in that it allows for others to access your data without creating redundant copies of files. We are particularly interested in using persistent identfiers (PIDs), which are URLs that do not change over time. One type of PID that we are very familiar with in research is a DOI (Digital Object Identifier). These are assigned to our publications, for example. 

There are three options to specify download URLs in this code:
1) List stable URLs in a "datalinks.txt" file in your "inputs" folder, and then iterate over these to download them. This option is ideal for smaller files that do not require user authentication to access.
2) Generate list of stable URLs using database intuition: this is ideal for ftp and http servers, where databasing syntax is easily inferred from the file URLs, without need for an API (advanced programming interface) to modify the search. Here, we can use regular expressions to guess the URLs for data representing different locations, points in time, or variables, or customize URLs to include details like our own username and password.
3) Accessing data though an API (Advanced Programming Interface): APIs allow us to use code to interact with database search parameters, find relavant data, and access download links for that relevant data. APIs help with searching complex, heterogenous geodatabases. For very large datasets (reanalysis data, satellite remote sensing data, etc), certain APIs allow us to subset data (say for a specific location) prior to downloading, reducing storage requirements. APIs often require environmnetal configuration for access control. If so, **access configuration details must be specified in README**

## Handling generated or simulated data
For experiments relying on synthetic data, please include synthetic data generation script in this notebook. Be sure to use features like set seed to ensure consistency with random number generation. Save all outputs to rawdata folder.


<span style='color:red'>delete before publishing</span>

# Define download file helper functions
This function will download files from URLs to the raw data folder created in the parent folder of the current repository. 
```{r helpers-for-download}
get_filename_from_url <- function(url) {
    httr2::url_parse(url)$path |> basename()
}
download_file <- function(url, out_dir) {
    old_o <-  options(timeout = max(300, getOption("timeout")))
    filename <- get_filename_from_url(url)

    filepath <- fs::path(out_dir, filename)
    if (fs::file_exists(filepath)) {
        print("File already exists, skipping: ", filename)
        options(old_o); return(TRUE)
    } else {
        print("Downloading: ", filename)
        req <- httr2::request(url) |>
            httr2::req_retry(max_tries=3, max_seconds=TIMEOUT, retry_on_failure=T,backoff=\(x){2}) |>
            httr2::req_progress()
        httr2::req_perform(req, path = filepath)
        print("✅ Saved: ", filepath)
        options(old_o); return(TRUE)
    }
}
```

<span style='color:red'>delete before publishing</span> 

# Example option 1: Downloading from a list of URLs in inputs/datalinks.txt
To use this option, copy and paste download urls of relevant datasets into the text folder located at inputs/datalinks.txt.

```{r}
# ===== USER SETTINGS =====
LINKS_FILE <- "inputs/auto_datalinks.txt"
OUTPUT_DIR <-  raw_data_dir
DELAY_SECONDS <- 2  # Check your data provider's API access instructions to see what delay requirements exist
TIMEOUT <- 30       # Seconds before request timeout, check your data provider's API access instructions to change as needed.
# =========================
```

```{r}
message("Downloading automatic datasets...")
auto_links <- readLines(LINKS_FILE)
if (length(auto_links) < 1L) {stop("No data links found in ",LINKS_FILE)}
for (url in auto_links) {
    download_file(url, OUTPUT_DIR)
    if (DELAY_SECONDS > 0) message("Waiting ",DELAY_SECONDS," s..."); Sys.sleep(DELAY_SECONDS)
}
message("🎉 Automatic datasets downloaded")
rm(auto_links)
```

```{r load-manual-data}
G <- data.table::fread(fs::path(raw_data_dir,"manual/G_reg.csv"), keepLeadingZeros = T)
stations <- data.table::fread(fs::path(raw_data_dir,"manual/asemat.csv"), keepLeadingZeros = T)
gwas <- data.table::fread(fs::path(raw_data_dir,"manual/pv_alueet.csv"), keepLeadingZeros = T)
pipes <- data.table::fread(fs::path(raw_data_dir,"manual/paikat.csv"), keepLeadingZeros = T)
```

<span style='color:red'>delete before publishing</span> 

# Option 2: using regular expressions to define URLs
    
## Decoding the file structure of online geodatabases
Just like we can use code to find and access files on our local machine, we can use code to find and access files on public geodatabases. Since these geodatabases are version controlled, providing code that links to the online files helps prevent us from making redundant copies of data on the internet. Programatically accessing public geodatabases requires that we understand how the database itself has been organized.

  - Click on the following link to the National Oceanic and Atmospheric Association databse website: https://psl.noaa.gov/data/gridded/
  - Navigate to the CEP/NCAR Reanalysis dataset"
  - Of the seven sections they've divided data into, click on "Surface"
  - Under "Air Temperature: Daily", click "See list"
  - Under "Surface", click "See list"

### TASK: Right click on the first link in the list, and select "copy link". Paste that link address below:
https://downloads.psl.noaa.gov/Datasets/ncep.reanalysis.dailyavgs/surface/air.sig995.1948.nc

There is anatomy to this link.
* https://downloads.psl.noaa.gov/Datasets: this is the stable location of many different types surface level (psl) meteorological data stored by NOAA.gov
* ncep.reanalsys.dailavgs : this gives us information about a specific dataset, here we are accessing the NCAR NCEP reanalysis data, daily averages
* air.sig995. : this tells us we are getting air tempearture at the near surface (indicated here by geopotential height of 995mBar pressure level
* 1948 : this is a year
* .nc : this is our file extension

In the link above, we can break the filepath down into substrings, using basic text commands:

```{r}
# ===== USER SETTINGS FOR SPI & SPEI DATA =====
OUTPUT_DIR <- fs::dir_create(raw_data_dir,"auto","ecmwfr_spi_spei_monthly")
TIMEOUT <- 1000 # Seconds before req timeout
# =========================
```

```{r download-spi-spei}
tryCatch(invisible(ecmwfr::wf_get_key()), error=\(e){stop("Add European Centre for Medium-Range Weather Forecasts API key (ecmwfr::wf_set_key())")}) # Check for key
# For more on ecmwfr datasets: ecmwfr::wf_dataset_info(), ecmwfr::wf_datasets()

ecmwfr_spi_spei_req_year <- function(y) {
    list(dataset_short_name = "derived-drought-historical-monthly",
        variable = c("standardised_precipitation_index",
                    "standardised_precipitation_evapotranspiration_index"),
        accumulation_period = c("1","3","6","12","24","36","48"),
        version = "1_0",
        product_type = "reanalysis",
        dataset_type = "consolidated_dataset",
        year = as.character(y),
        month = c("01","02","03","04","05","06","07","08","09","10","11","12"),
        area = c(70, 18, 59, 32), # Nmax, Emin, Nmin, Emax, WGS84
        target = paste0("ecmwfr_spi_spei_",y,".nc")  # Output filename & format
    )
}
# test_dl <- ecmwfr::wf_request(ecmwfr_spi_spei_req_year(1966), path=fs::dir_create(raw_data_dir,"auto","test"), time_out = 1000)

message("Downloading monthly SPI and SPEI...")
message("This can take around half an hour")
ecmwfr::wf_request_batch(
    request_list = lapply(1965:2024, ecmwfr_spi_spei_req_year),
    workers = 20, # Max 20?
    time_out = TIMEOUT, #s
    path = OUTPUT_DIR
)
# Unzip all downloads into parent folder, delete zips
dl_zips <- fs::dir_ls(OUTPUT_DIR, type="file", glob="*.zip")
stopifnot(all(fs::is_file(dl_zips)))
for (f in dl_zips) { unzip(f, exdir = fs::path_dir(f));  unlink(f) }
rm(dl_zips)

# Possible to download also qual criteria for SPI & SPEI
# Similar req to above, but variables "probability_of_zero_precipitation_spi", "test_for_normality_spi", "test_for_normality_spei"
# All three cover all years in the dataset (no year needed in req)

message("SPI and SPEI datasets downloaded.")
```

```{r parse-spi-spei-files}
# Extract variables and dates from SPI & SPEI filenames
spi_spei_nc_files <- fs::dir_ls(OUTPUT_DIR, type="file",glob="*.nc")

# Assuming format: "variable_(infos)_YYYYMM.(geo.infos).nc"
file_meta <- tibble::tibble(filepath = as.character(spi_spei_nc_files)) |> 
  dplyr::mutate(
    filename = fs::path_file(filepath), # Next line: drop the tail infos with "." sep
    filename_cleaned = stringr::str_split_i(fs::path_ext_remove(filename), "\\.", i=1) 
  ) |> 
  tidyr::separate(filename_cleaned, sep = "_",  # Split at the underscore
    into = c("variable",NA,NA,NA,NA,NA,"time_str")) |> #NAs: distrib,spat_ext,reanalysis,output_stream,ref_per
  dplyr::mutate(year = as.integer(stringr::str_sub(time_str, 1, 4)),
                month = as.integer(stringr::str_sub(time_str, 5, 6))
  ) |> data.table::as.data.table()
```

```{r}
# Get a spatial object of station coordinates
station_coords <- sf::st_as_sf(stations[,.(station_id=tunnus,lon,lat)],
                                coords=c("lon","lat"), crs="EPSG:3067") |> # Original crs ETRS-TM35FIN
    sf::st_transform("EPSG:4326") # to WGS84
```

```{r build-spi-spei-timeseries}
# A function to process a single variable efficiently
extract_timeseries <- function(var_name, files_df, pts_sf) {
    var_files <- files_df[variable == var_name][order(time_str)] # Files for only this var, sort chronologically

    raster_stack <- terra::rast(var_files$filepath) # Create SpatRaster stack. terra handles virtually, saving mem

    # Ensure CRS of supplied points and read NetCDF files match
    if (sf::st_crs(pts_sf)$input != sf::st_crs(raster_stack)$input) {
        message("Converting points' CRS (",sf::st_crs(pts_sf)$input,") to file's CRS (",sf::st_crs(raster_stack)$input,")")
        pts_sf <- sf::st_transform(pts_sf, sf::st_crs(raster_stack)) # Transform the points to match files
    }
    # Convert sf points to terra's SpatVector format
    pts_vect <- terra::vect(pts_sf)
    browser()
    # Ensure that SpatRaster layers are chronological (may be default beavior, but JIC)
    raster_stack <- raster_stack[[order(terra::time(raster_stack))]]
    # Extract data
    extracted <- terra::extract(raster_stack, pts_vect, fun=identity, bind=TRUE, exact=TRUE, ID=FALSE)

    # Combine with site IDs and tidy into a long format
    res_long <- tidyr::pivot_longer(
        data = as.data.frame(extracted), # df rows=points, cols = time layers (var,var.1,var.2 ...)
        cols = -station_id,
        names_to="layer_name", values_to="value"
    ) |> 
    # Map the chronological time back onto the extracted layers
    dplyr::group_by(station_id) |> 
    dplyr::mutate(variable = var_name,
        time_str=var_files$time_str, year=var_files$year, month=var_files$month
    ) |> dplyr::ungroup() |> 
    dplyr::select(station_id, time_str,year,month, variable, value)

    return(res_long)
}

# Extract over each variable (SPI1, SPI3, ..., SPEI36, SPEI48) and bind into one df
station_spi_spei_ts <- purrr::map_dfr(unique(file_meta$variable), 
  ~extract_timeseries(.x, file_meta, station_coords)
)
```

Download FMI's daily gridded data from around all pipes where natural QC'd GW data exists.
Data DOI reference: https://doi.org/10.35614/ISSN-2341-6408-IK-2022-11-RL
AWS S3 link assumed to be stable enough
```{r}
BUFF_M <- 200
areas_around_pipes <- pipes[paikka_id %in% G$paikka_id][,.(lon_min=lon-BUFF_M, lon_max=lon+BUFF_M,lat_min=lat-BUFF_M, lat_max=lat+BUFF_M), by=paikka_id]
metR::GlanceNetCDF("https://fmi-gridded-obs-daily-1km.s3-eu-west-1.amazonaws.com/Netcdf/Tday/tday_1978.nc",
    subset = list(lon=areas_around_pipes[1,seq(lon_min,lon_max)],
                  lat=areas_around_pipes[1,seq(lat_min,lat_max)])
)
test_r <- terra::rast("C:/Users/ppoykko19/Koodi/waterdigi-drought-pred/inputs/manual/monthly_19610101T000000_19610101T000000_20101201T000000_latlon.grb2")
```

```{r}
test_r <- terra::rast("C:/Users/ppoykko19/Koodi/waterdigi-drought-pred/inputs/manual/scPDSI_1901-2024_global_ cru_ts4.09/scPDSI.cru_ts4.09_1901-2024.bams.2025.GLOBAL.IGBP.WHC.nc")
summary(test_r)

plot(test_r)
```

<span style='color:red'>delete before publishing</span> 
# Option 3: generate list of filepaths using API for an online geodatabase
#TODO: Update with AquaINFRA

There are dozens of rich online geodatabases for governmental, public, and research data available. Most of this geodatabases will provide both a graphical user interface (GUI) and advanced programming interface (API) to search, select, and download their data holdings. The code below was generated using the Copernicus Climate Data Store (CDS) web-based GUI. This graphical user interface returns a scripted python API call that I can copy into my notebook to download my data. To generate this code, I went to the following website:

(https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels-monthly-means?tab=download)

And filled out the form to specify my search parameters. I then went to the bottom of the page, and under "Corresponding API Code", seleceted "Show API Request Code"

Please note that I must download the cdsapi python package to execute this request, and indicate my Copernicus login credentials by modifying the $HOME/.cdsapirc file created by this python library for this to work.
[How to configure the Copernicus Climate Data Store API on your machine](https://cds.climate.copernicus.eu/how-to-api)

Learn more about geodatabase authentication protocols on the "Data access authentication" notebook in the how-to page on https://github.com/DigitalWaters-fi.

```{python}
 import cdsapi
    
    dataset = "reanalysis-era5-single-levels-monthly-means"
    request = {
        "product_type": ["monthly_averaged_reanalysis"],
        "variable": [
            "2m_temperature",
            "total_precipitation",
            "clear_sky_direct_solar_radiation_at_surface",
            "instantaneous_surface_sensible_heat_flux",
            "surface_latent_heat_flux",
            "runoff"
        ],
        "year": [
            "2000", "2001", "2002",
            "2003", "2004", "2005",
            "2006", "2007", "2008",
            "2009", "2010", "2011",
            "2012", "2013", "2014",
            "2015", "2016", "2017",
            "2018", "2019", "2020"
        ],
        "month": [
            "01", "02", "03",
            "04", "05", "06",
            "07", "08", "09",
            "10", "11", "12"
        ],
        "time": ["00:00"],
        "data_format": "grib",
        "download_format": "unarchived",
        "area": [72, 19, 58, 32]
    }
    
    client = cdsapi.Client()
    client.retrieve(dataset, request).download()
```

## Need help?

Search, ask, and answer data access questions at https://github.com/orgs/DigitalWaters-fi/discussions

Tag #dataaccess

