---
title: "Standard Element 1"
subtitle: "Download source data"
author: "Pietari Pöykkö"
date: last-modified
format: html
---

The Finnish natural-state groundwater data central in this analysis is not available
as one contained dataset. This data has not been productized. Instead, data must
be downloaded manually through the Hertta open data portal of Syke, with a manual
selection of stations. Link to this portal is provided in README. Additionally,
I have trimmed down the data from natural-state stations to only contain maximally
long time series. Moreover, I have quality-controlled this dataset. The QC
procedure is described in Pöykkö et al. 2026.
I have not asked for a permission to share this customized dataset in a data repository.

Running this notebook additionally requires an API key for European Centre for Medium-Range Weather Forecasts services.
This key can be obtained by registering an ECMWF account on <https://www.ecmwf.int/>, and viewing the key from your personal profile.
The script will ask for this key automatically, or it can be provided by running the command `ecmwfr::wf_set_key()`

The raw data is stored in the project subfolders of `inputs`.
These are not committed to the repository, using .gitignore.


## Data description
Describe your data sources here. Reference the table provided on your README.

## How to run this notebook
Provide a detailed description of inputs, input directories, outputs, and output directories.


## Storage requirements
The total storage required for running this notebook is 1.1 Gb.
 - 330 Mb will be downloaded
 - 650 Mb for the unzipped downloads
 - 50 Mb for manually accessed GW data
Largest individual downloaded file is around 5 Mb.
Largest of all files is the main GW datafile at 13 Mb.

### Make sure you have sufficient storage available to download all data required
```{r}
if ((ps::ps_disk_usage()$available / 2^30) < 2) { # Gb
    stop("Insufficient disk space for the automatically downloaded datasets")
}
```

```{r}
#| label: "common"
#| include: FALSE
source("_common.R") # Knitr hook to enable global R-env when rendering many .qmd
```

```{r init}
here::i_am("SE1_data_access.qmd")
root = here::here()

raw_data_dir = fs::path(root, "inputs") # Define raw data dir var

fs::dir_create(raw_data_dir, c("manual","auto")) # Ensure folder structure

if (length(fs::dir_ls(fs::path(raw_data_dir, "manual"))) < 4L) { # TODO: set precise number of required manual dls
    stop("Datasets listed in README to require manual inputs are not found")
}
```

## Define download file helper functions
These functions will download files from URLs to the raw data folder created in the current repository. 
```{r helpers-for-download}
get_filename_from_url <- function(url) {
    httr2::url_parse(url)$path |> basename()
}
download_file <- function(url, out_dir, t_out) {
    old_o <-  options(timeout = max(300, getOption("timeout"))) # why this is needed...
    on.exit({options(old_o)}, add=TRUE)

    filename <- get_filename_from_url(url)

    filepath <- fs::path(out_dir, filename)
    if (fs::file_exists(filepath)) {
        message("File already exists, skipping: ", filename)
        return(TRUE)
    } else {
        message("Downloading: ", filename)
        req <- httr2::request(url) |>
            httr2::req_retry(max_tries=3, max_seconds=t_out,
                retry_on_failure=TRUE,backoff=\(x){2}) |>
            httr2::req_progress()
        httr2::req_perform(req, path = filepath)
        message("✅ Saved: ", filepath)
        return(TRUE)
    }
}
```

## Read in manually accessed GW data
```{r load-manual-data}
G <- data.table::fread(fs::path(raw_data_dir,"manual/G_reg.csv"), keepLeadingZeros = T)
stations <- data.table::fread(fs::path(raw_data_dir,"manual/asemat.csv"), keepLeadingZeros = T)
gwas <- data.table::fread(fs::path(raw_data_dir,"manual/pv_alueet.csv"), keepLeadingZeros = T)
pipes <- data.table::fread(fs::path(raw_data_dir,"manual/paikat.csv"), keepLeadingZeros = T)
```


## Download SPI & SPEI timeseries data
```{r}
# ===== USER SETTINGS FOR SPI & SPEI DATA =====
OUTPUT_DIR <- fs::dir_create(raw_data_dir,"auto","ecmwfr_spi_spei_monthly")
TIMEOUT <- 2000 # Seconds before req timeout
# =============================================
```

```{r download-spi-spei}
tryCatch(invisible(ecmwfr::wf_get_key()), # Check for ECMWFR API key
    error=\(e){ warning("Add ECMWF API key (ecmwfr::wf_set_key())")
                ecmwfr::wf_set_key()
              }
)
# For more on ecmwfr datasets: ecmwfr::wf_dataset_info(), ecmwfr::wf_datasets()

ecmwf_spi_spei_req_year <- function(y) {
    list(dataset_short_name = "derived-drought-historical-monthly",
        variable = c("standardised_precipitation_index",
                    "standardised_precipitation_evapotranspiration_index"),
        accumulation_period = c("1","3","6","12","24","36","48"),
        version = "1_0", product_type = "reanalysis",
        dataset_type = "consolidated_dataset",
        year = as.character(y),
        month = c("01","02","03","04","05","06","07","08","09","10","11","12"),
        area = c(70, 18, 59, 32), # Nmax, Emin, Nmin, Emax, WGS84
        target = paste0("ecmwfr_spi_spei_",y,".nc")  # Output filename & format
        # ("target" potentially converts to zip if the req would return many files)
    )
}
# test_dl <- ecmwfr::wf_request(ecmwf_spi_spei_req_year(1966), path=fs::dir_create(raw_data_dir,"auto","test"), time_out = 1000)

out_dir_dls <- fs::dir_create(OUTPUT_DIR, "downloads") # Dir for dls from API
out_dir_final <- fs::dir_create(OUTPUT_DIR, "unzipped") # Dir for final usable files

all_ecmwf_reqs <- lapply(1965:2024, ecmwf_spi_spei_req_year)

# Drop the API calls which files already exists
wanted_req_files <- fs::path_ext_remove(sapply(all_ecmwf_reqs, \(req) req$target))
existing_req_files <- fs::path_ext_remove(fs::path_file(fs::dir_ls(out_dir_dls, type="file")))
needed_ecmwf_reqs <- all_ecmwf_reqs[which(!(wanted_req_files %in% existing_req_files))]

if (!rlang::is_empty(needed_ecmwf_reqs)) {
    message("Downloading monthly SPI and SPEI...")
    message("This can take around half an hour")
    sent_reqs <- ecmwfr::wf_request_batch(
        request_list = needed_ecmwf_reqs,
        workers = min(20, length(needed_ecmwf_reqs)), # Max 20?
        time_out = TIMEOUT, #s
        retry = 20, # sec to wait between api calls
        path = out_dir_dls # Into subfolder of downloads
    )
    # Unzip all downloads into subfolder
    all_dl_zips <- fs::dir_ls(out_dir_dls, type="file", glob="*.zip")
    stopifnot(all(fs::is_file(all_dl_zips)))
    for (f in all_dl_zips) { unzip(f, exdir = out_dir_final) }
    rm(all_dl_zips, sent_reqs)
    # Possible to download also quality criteria for SPI & SPEI
    # Similar req to above, but variables "probability_of_zero_precipitation_spi", "test_for_normality_spi", "test_for_normality_spei"
    # All three cover all years in the dataset (no year needed in req)
    message("SPI and SPEI datasets downloaded.")
} else { message("All requested SPI & SPEI original files found") }
rm(wanted_req_files, existing_req_files, needed_ecmwf_reqs, all_ecmwf_reqs, out_dir_dls)
```

```{r finish}
message("Finished")
```

Download FMI's daily gridded data from around all pipes where natural QC'd GW data exists.
Data DOI reference: https://doi.org/10.35614/ISSN-2341-6408-IK-2022-11-RL
AWS S3 link assumed to be stable enough
[DRAFT, likely will be deleted]
