---
title: "Standard Element 3"
subtitle: "Model training, validation, and interpretation"
author: "Pietari Pöykkö"
date: last-modified
format: html
---

This submission includes:

Python or R code for training a statitical model
Python or R code to extract relevant information from model output
Python or R code to calculate model fit coefficients.
README is edited detailing workflow of model train code, including locations of inputs and outputs in your project directory.
Note: having a plan for your figures will make your life easier for deciding what relevant information you need to plan for model output!

Be sure to edit your Data and methods document with any updates from this stage of analysis, and get started writing your results and discussion!

Reference for best practices for model development: https://journals.ametsoc.org/view/journals/aies/2/2/AIES-D-22-0065.1.xml 

```{r}
#| label: "common"
#| include: FALSE
if(!interactive()) {
  source("_common.R") # Knitr hook to enable global R-env when rendering many .qmd
}
```

```{r drought-index-correlation}
# Function to re-format cor.test output into 1-long lists (output will stay as 1 row when applying in dt)
dt_cor.test <- function(...) {
  res_list <- as.list(stats::cor.test(...))
  res_list[["conf.int.l"]] <- res_list[["conf.int"]][[1]]
  res_list[["conf.int.u"]] <- res_list[["conf.int"]][[2]]
  res_list[["conf.int"]] <- NULL;  stopifnot(lapply(res_list, length) == 1L)
  return(res_list)
}
pipe_correlations <- pipe_sgi_w_dr_inds[, dt_cor.test(sgi, value, method="pear"), by=list(paikka_id, dr_ind)]

# Plots
pipe_correlations |> 
  dplyr::mutate(dr_ind = factor(dr_ind, levels = gtools::mixedsort(unique(dr_ind)))) |> 
  ggplot(aes(dr_ind, estimate, color=dr_ind)) +
      geom_violin() +
      ggforce::geom_sina(aes(alpha=p.value <= 0.01, group=dr_ind)) +
      stat_summary(fun.data = "mean_se", color="black")
      # geom_point(aes(alpha=p.value <= 0.01))


station_sgi_w_dr_inds <- merge(station_sgi_long, station_spi_spei_ts, by=c("station_id","time"))
station_correlations <- station_sgi_w_dr_inds[, dt_cor.test(sgi, value, method="pear"), by=list(station_id, dr_ind)]
station_correlations |> 
  dplyr::mutate(dr_ind = factor(dr_ind, levels = gtools::mixedsort(unique(dr_ind)))) |> 
  ggplot(aes(dr_ind, estimate, color=dr_ind)) +
      geom_violin() +
      ggforce::geom_sina(aes(alpha=p.value <= 0.01, group=dr_ind)) +
      stat_summary(fun.data = "mean_se", color="black")
      # geom_point(aes(alpha=p.value <= 0.01))


acf(pipe_sgi_long[paikka_id == 10550, sgi], na.action = na.pass)
acf(pipe_sgi_long[paikka_id == 10550, sgi], na.action = na.exclude)

ccf_test_spei <- station_spi_spei_ts[station_id == pipes[paikka_id == 10550, asema_tunnus] &
  variable == "SPEI12"
]
ccf_test_pipe <- pipe_sgi_long[paikka_id == 10550]
ccf_test <- merge(ccf_test_spei, ccf_test_pipe, by="dr_ind")

ccf(ccf_test$sgi, ccf_test$value, na.action = na.pass)
stats::cor.test(ccf_test$sgi, ccf_test$value,  method="pearson")

plot(ccf_test$sgi, ccf_test$value)

```

```{r}

DescTools::RMSE()
```



Test out buffering around pipes
```{r}
BUFF_M <- 200
areas_around_pipes <- pipes[paikka_id %in% G$paikka_id][,.(lon_min=lon-BUFF_M, lon_max=lon+BUFF_M,lat_min=lat-BUFF_M, lat_max=lat+BUFF_M), by=paikka_id]
```


